---
title: "Factors Lab"
format: html
embed-resources: true
---

## Collapsing factors

- Load covid testing line list data 

```{r echo = TRUE, message = FALSE, warning=FALSE}
# install.packages("medicaldata")
library(medicaldata)
library(tidyverse)
covid_data <- medicaldata::covid_testing
```

- make `patient_class` a factor using `as.factor()` function. How many levels does it have?


- count appearances of each level in the data.

- create a new `patient_class2` column/variable by collapsing 9 levels of the `patient_class` factor into 5 levels by grouping some of them together. Count appearances of these new levels in the data

## 

```{r echo = TRUE, message = FALSE}
covid_data <- covid_data |> 
  mutate(patient_class = as.factor(patient_class))

covid_data |> 
  count(patient_class)
```

##

```{r echo = TRUE, message = FALSE}
covid_data <- covid_data |> 
  mutate(patient_class2 = 
           fct_collapse(patient_class,
                        inpatient = c("inpatient"),
                        surgery = c("admit after surgery-ip", "admit after surgery-obs", "day surgery"),
                        outpatient = c("outpatient", "recurring outpatient", "observation"),
                        emergency = c("emergency"),
                        not_applicable = c("not applicable")
                        ))

covid_data |> 
  count(patient_class2)
```
