---
title: "Day 4 Lab Data Wrangling and Viz"
format: 
  html:
    toc: true
    toc-expand: true
    toc-depth: 2
    
editor: visual
execute:
  echo: fenced
theme: cerulean
embed-resources: true
---

```{r}
library(tidyverse)
```

# NYC Citi Bikes Dataset

The dataset can be found in the data folder. However, take a few minutes to visit the Citi-Bikes website (https://citibikenyc.com/system-data) to learn about their data.



# Examining our data

## Question 1

Begin by reading in the data. Use `glimpse()` to get a preview of the data and view the columns.

```{r}
glimpse(citi)
```

Clean the names so that they are consistent with `tidyverse` conventions.


How many rows and columns does the dataset have? What do they represent?

## Question 2

Let's now learn something about the data.

a. Can you tell which time period this dataset covers?


b.  What are the most popular pick-up locations across the city for NY Citi Bike rental during this time period? What is the post popular drop-off location?

c.  Which age group rents the most bikes?


d.  What is the longest trip recorder? Where did this user start from and end to?


## Question 3

Let's learn some information about this data. a. What are the most popular pick-up locations across the city for NY Citi Bike rental? (see before)

b.  Does this change between depending on whether the trip happens on a weekday or a weekend?

Hint: to figure this out, you will first need to create a new column using `mutate()`. This will need to classify a day in `weekday` or `weekend`.


## Question 4

Let's look only at trips that happen in the weekend. How does the average trip duration vary across different age groups in these days?

Do this using the `filter()`, `group_by()`, and `summarise()` functions.


In the weekdays :


::: callout-note
## Trouble with filter

There are some other packages with functions named `filter()` or `select()` and if those packages are loaded most recently then sometimes you can get problems. In the future when you are doing data cleaning if `filter()` or `select()` are not working but you are confident you have called them correctly, check to make sure `tidyverse` is your most recently loaded package.
:::

# Data Visualization

Using the skills you have learned during lecture, try making plots that answer the following questions.

## Question 6

How does bike rental vary across the two user groups (one-time users vs subscribers) on different days of the week?



## Question 7 (Extra exercises for practice)

a.  Create a bar plot illustrating the number of trips that start at each of the 10 most popular pick up locations. Make sure to include titles, labels, and to make the colors clear.


b.  Earlier, you found the average trip length from the original station to the top three drop off - now we are interested instead in visualizing the distribution of these trip lengths. As always, there are many ways you can represent this! Try out of few, with boxplots, histograms, or density plots (`geom_density`) Note , you can filter out outliers if it will make your plot more understandable.


