---
title: "Day 11 Lab: Linear Mixed Effects Model"
format: 
  html:
    toc: true
    toc-expand: true
    toc-depth: 2
editor: visual
embed-resources: true
execute:
  echo: fenced
theme: cerulean
---

In this chapter, we'll be working with some real data from a study looking at the effects of sleep deprivation on psychomotor performance (Belenky et al., 2003)\[\^1\]. Data from this study is included as the built-in dataset `sleepstudy` in the `lme4` package for `R`.

# Understanding your data

## Question 1

First we need to load in the data.

```{r}
set.seed(72024)

library(tidyverse)
library(lme4)

# Use data() to load a data set from a package into your environment
data("sleepstudy")
```

Examine your data. Use `?sleepstudy` to look at the documentation for this dataset.

## Question 2

Notice the documentation says "days 0-1 were adaptation and training (T1/T2), day 2 was baseline (B); sleep deprivation started after day 2". For our purposes we don't care about the days they were training participants, we want to look at the days they were sleep deprived.

Run the code below to make a new variable `days_since_baseline` and remove observations from days 0 and 1. Notice we are making a new data frame instead of overwriting the one from the package.

```{r}
sleep_study_data <- sleepstudy |> 
  mutate(days_since_baseline = Days - 2) |> 
  filter(Days >= 2)
```

Now let's look at the data for a single subject to better understand our data.

```{r}
subject308_data <- filter(sleep_study_data, Subject == "308")

ggplot(subject308_data, aes(x = days_since_baseline, y = Reaction)) +
  geom_point() +
  geom_line(linetype = "dashed") +
  theme_bw()
```

Notice there are only 18 people in this study (can run `unique(sleep_study_data$Subject)`). We can reasonably plot all of this data at once.

```{r}
ggplot(sleep_study_data, aes(x = days_since_baseline, y = Reaction)) +
  geom_point() +
  geom_line(linetype = "dashed") +
  theme_bw() +
  facet_wrap(~Subject)
```

What general patter(s) do you see in the data? Would it be reasonable to assume independence between data points?

# Modeling

For our data exploration purposes let's consider three different models:

1.  linear model (we know this is incorrect because it breaks the independence assumption, but let's see what happens...)
2.  linear mixed effects models with subject-specific intercepts
3.  linear mixed effects models with subject-specific intercepts and slopes

## Question 3

Write out the theoretical regression equation for each proposed model.

## Question 4

Fit model 1 and save it into an object called `model1`.

After the model is fit use the following code to examine how well our linear regression equation represents the data. Does it do a good job?

```{r}
#| eval: false

ggplot(sleep_study_data, aes(x = days_since_baseline, y = Reaction)) +
  geom_abline(
    intercept = model1$coefficients["(Intercept)"],
    slope = model1$coefficients["days_since_baseline"],
    color = 'blue'
  ) +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~ Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

## Question 5

Fit proposed model 2 and save it into an object called `model2`.

After the model is fit use the following code to examine how well our linear regression equation represents the data. Does it do a good job?

```{r}
#| eval: false

# Make new data frame with predictors
model2_prediction_data <- crossing(
  Subject = sleep_study_data %>% pull(Subject) %>% levels() %>% factor(),
  days_since_baseline = 0:7
) 

# Add column of predicted values
model2_prediction_data <- mutate(
  model2_prediction_data,
  Reaction = predict(model2, model2_prediction_data)
)

# Plot our regression lines for all participants
ggplot(sleep_study_data, aes(x = days_since_baseline, y = Reaction)) +
  geom_line(
    data = model2_prediction_data,
    color = 'blue'
  ) +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~ Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

## Question 6

Fit proposed model 3 and save it into an object called `model3`.

After the model is fit use the following code to examine how well our linear regression equation represents the data. Does it do a good job?

```{r}
#| eval: false

# Make new data frame with predictors
model3_prediction_data <- crossing(
  Subject = sleep_study_data %>% pull(Subject) %>% levels() %>% factor(),
  days_since_baseline = 0:7
) 

# Add column of predicted values
model3_prediction_data <- mutate(
  model3_prediction_data,
  Reaction = predict(model3, model3_prediction_data)
)

# Plot our regression lines for all participants
ggplot(sleep_study_data, aes(x = days_since_baseline, y = Reaction)) +
  geom_line(
    data = model3_prediction_data,
    color = 'blue'
  ) +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~ Subject) +
  labs(y = "Reaction Time", x = "Days deprived of sleep (0 = baseline)")
```

# Interpretation

Ultimately, we want to investigate the linear relationship between reaction time and sleep deprivation. Hopefully at this point you have found the model with random intercepts and slopes to best represent the data.

Let's look at the fixed effect for sleep deprivation days (`days_since_baseline`).

For all the three models, what is the estimate of the association between sleep deprivation and reaction time?

```{r}
summarize_ci_lr <- function(model,name_var,level=0.95){
  est <- summary(model)$coefficients[name_var,"Estimate"]
  lb <- confint(model, level=level)[name_var, 1]
  ub <- confint(model,level=level)[name_var, 2]
  
  lb_name <- paste((1-level)/2*100, "%")
  ub_name <- paste((1-(1-level)/2)*100, "%")
  est_ci <- c(est,lb,ub)
  names(est_ci) <- c("Estimate",lb_name, ub_name)
  return(est_ci)
}

summarize_ci_mixed_effects <- function(model,name_var, level=0.95){
  est <- fixef(model)[name_var]
  lb <- confint(model, level=level,quiet=T)[name_var, 1]
  ub <- confint(model,level=level,quiet=T)[name_var, 2]
  
  lb_name <- paste((1-level)/2*100, "%")
  ub_name <- paste((1-(1-level)/2)*100, "%")
  est_ci <- c(est,lb,ub)
  names(est_ci) <- c("Estimate",lb_name, ub_name)
  return(est_ci)
  
}

```

Summary for model 1

```{r}
summarize_ci_lr(model1,"days_since_baseline")
```

```{r}
summarize_ci_mixed_effects(model2,"days_since_baseline")

```

```{r}
summarize_ci_mixed_effects(model3,"days_since_baseline")
```

*For the average individual*, we estimate an increase in average reaction time of `r fixef(model3)[2]`ms to be associated with a 1 day increase in sleep deprivation.

Working with the `lme4` objects... to get random effects and fixed effects use:

```{r}

ranef(model3)
fixef(model3)


```

[^1][Belenky, G., Wesensten, N. J., Thorne, D. R., Thomas, M. L., Sing, H. C., Redmond, D. P., Russo, M. B., & Balkin, T. J. (2003). Patterns of performance degradation and restoration during sleep restriction and subsequent recovery: A sleep dose-response study. Journal of Sleep Research, 12(1), 1–12.](https://onlinelibrary.wiley.com/doi/full/10.1046/j.1365-2869.2003.00337.x)

[^1]: 1

<!-- Exercises derived from : <https://psyteachr.github.io/stat-models-v1/introducing-linear-mixed-effects-models.html> -->
