---
title: "Linear mixed effects model: R example"
format: html
editor: visual
embed-resources: true
---

## When to use mixed effects models:

- Independence assumptions will sometimes (often!) break down.

  - Think for instance of being interested in relationship between rain and malaria, and collecting data at village levels.
  - Or think of asssessing success of a treatment and collecting data at hospital level.

- We use mixed effects models to allow us to correctly estimate the uncertainty in our model when independence does not hold.

- Remember that in this your "true" sample size is smaller, so uncertainty will be larger.

- Mixed effects models let you control for the variables of interested without coding each one as a categorical variable (what would be the issue with doing that?)

- We can specify both a model with **random slope** or **random intercept** (or both).

### Fixed vs Random Effects :

**Fixed Effect** :\
- the variables we want to answer research questions about

\- variables that researchers would care about if we were to collect data to replicate the study in a new location/group.

**Random effects** :

\- less related to the scientific question of interest , but we must account for them to address the relationship of interest

\- effects where we do not include all levels of the category.

## Interpretation

You can interpret the slopes as :

1\) Changes in population average.

2\) Changes for "typical" individuals, i.e. $\beta_1$ is the expected change in the response for subpopulations that differ by 1 unit for individuals for which the $b_1=0$.

(Similar for the intercept)

## Generalizing from linear model..

- You can also fit mixed effects models with a non-linear link - so fit GLMs with random effects to adjust for break down of independence assumption.
- However, interpretation here because more difficult, as we are always looking at the conditional effects (individual specific).

### Alternatives?

Another type of model that can be appropriate for correlated data are Generalized Estimating Equation Models (GEE).

## R example

```{r}
library(tidyverse)
library(lme4)
```

## data {#data}

```{r}
chart_dat <- read_csv("data/chart_wide_condense.csv")
head(chart_dat)
dim(chart_dat)
```

## Question

1.  Do we see an average increase in the average Math Average Score between the years 2008 and 2010 for sixth grade students?
2.  Is such an increase different for charter vs non-charter school

### Data Organization

Key variables in `chart_wide_condense.csv` which we will examine to address the research question above are:

- `schoolid` = includes district type, district number, and school number
- `schoolName` = name of school
- `urban` = is the school in an urban (1) or rural (0) location?
- `charter` = is the school a charter school (1) or a non-charter public school (0)?
- `schPctnonw` = proportion of non-white students in a school (based on 2010 figures)
- `schPctsped` = proportion of special education students in a school (based on 2010 figures)
- `schPctfree` = proportion of students who receive free or reduced lunches in a school (based on 2010 figures). This serves as a measure of poverty among school families.
- `MathAvgScore.0` = average MCA-II math score for all sixth grade students in a school in 2008
- `MathAvgScore.1` = average MCA-II math score for all sixth grade students in a school in 2009
- `MathAvgScore.2` = average MCA-II math score for all sixth grade students in a school in 2010

This data is stored in WIDE format, with one row per school, as illustrated in Table \@ref(tab:table1chp9).

(Data and some code sourced from Beyond Multiple Linear Regression, data originally from Minnesota Department of Education (2018))

```{r, include=FALSE, warning=FALSE}
chart_long <- chart_dat %>%
  gather(key = "key", value = "MathAvgScore",
         MathAvgScore.0:MathAvgScore.2) %>%
  separate(key, into = c("name", "year08"), sep = "\\.") %>%
  dplyr::select(-c("name")) %>%
  arrange(schoolid, year08) %>%
  mutate(year08 = as.numeric(year08))

head(chart_long)
dim(chart_long)
```

## Fitting a simple model of time with a random intercept for school

Writing out the equation for this model:

For Year $i$ and School $j$:

$$MathAvgScore_{ij} | Year_{ij} = \beta_0 + b_{0j} + \beta_1 Year_i  + \epsilon_{ij}$$

```{r}
model1 <- lmer(MathAvgScore ~ year08 + (1|schoolid), 
  REML=T, 
  data= chart_long)
summary(model1)
```

# plot the fitted values

Lets look at the predicted values fro three schools:

```{r}


predict_dat = chart_long %>%
              filter(schoolid %in% c(
                                     "Dtype 7 Dnum 4008 Snum 10",
                                     "Dtype 1 Dnum 23 Snum 10",
                                     "Dtype 1 Dnum 2134 Snum 10")) 
predict_vals = predict(model1, 
                       predict_dat %>% dplyr::select(schoolid, year08))

predict_dat = cbind(predict_dat, predict_vals)  
plot_fit = predict_dat %>% 
            ggplot(aes(x = year08, y = predict_vals, color = schoolid)) +
            geom_line() +
            geom_point(aes(x = year08, y = MathAvgScore, color = schoolid)) +
            xlab("Years from 2008") +
            theme_bw()
            
plot_fit
```

## Fitting a model of time with a random intercept and random slope for school

Writing the equation for this model:

For Year $i$ and School $j$:

$$MathAvgScore_{ij} | Year_{ij} = \beta_0 + b_{0j} + \beta_1 Year_i + b_{1j} Year_i + \epsilon_{ij}$$

```{r}
model2 <- lmer(MathAvgScore~ year08 + (1 + year08|schoolid), 
  REML=T, data= chart_long)
summary(model2)
```

## Visualizing fitted values

```{r}

predict_dat2 = chart_long %>%
              filter(schoolid %in% c(
                                     "Dtype 7 Dnum 4008 Snum 10",
                                     "Dtype 1 Dnum 23 Snum 10",
                                     "Dtype 1 Dnum 2134 Snum 10")) 

predict_vals2 = predict(model2, predict_dat2 %>% dplyr::select(schoolid, year08))

predict_dat2 = cbind(predict_dat2, predict_vals2)  
plot_fit2 = predict_dat2 %>% 
            ggplot(aes(x = year08, y = predict_vals2, color = schoolid)) +
            geom_line() +
            geom_point(aes(x = year08, y = MathAvgScore, color = schoolid))+
            theme_bw()
            
plot_fit2

```

## Adding in charter school as a covariate

And interaction between `charter` and year.

For Year $i$ and School $j$:

$$MathAvgScore_{ij} | Year_{ij} = \beta_0 + b_{0j} + \beta_1 Year_i + b_{1j} Year_i + \beta_2 \mathbb{1}(Charter_{j}) +  \beta_3\mathbb{1}(Charter_{j})\times Year_i + \epsilon_{ij}$$

```{r}
model3 <- lmer(MathAvgScore~ charter*year08  + (year08|schoolid), 
  REML=T, data=chart_long)
summary(model3)
```

## Visualizing fitted values

```{r}

predict_dat3 = chart_long %>%
              filter(schoolid %in% c(
                                     "Dtype 7 Dnum 4008 Snum 10",
                                     "Dtype 1 Dnum 23 Snum 10",
                                     "Dtype 1 Dnum 2134 Snum 10",
                                     "Dtype 1 Dnum 500 Snum 50",
                                     "Dtype 7 Dnum 4188 Snum 10")) 

predict_vals3 = predict(model3, predict_dat3 %>% dplyr::select(schoolid, charter,year08))

predict_dat3 = cbind(predict_dat3, predict_vals3)  
plot_fit3 = predict_dat3 %>% 
            ggplot(aes(x = year08, y = predict_vals3, color = schoolid)) +
            geom_line(aes(linetype=as.factor(charter))) +
            geom_point(aes(x = year08, y = MathAvgScore, color = schoolid))+
            scale_linetype_discrete(name="Charter School:")+
            theme_bw()
            
plot_fit3

```

## How to get estimates for $b_{0j}$ and $b_{1j}$?

Each row for the specific `schoolid` :

```{r}
school_random_effects <- ranef(model3)
head(school_random_effects)

```

Remember that to get the specific intercept or slope for a certain school you will need to add these two $\beta_0$ and $\beta_1$.
