---
title: "Day 1 Lab"
format: html
editor: visual
---

This Quarto document operates similar to a word document with extra capabilities.

## Rendering your document

To produce an html document with this work, click the "Render" button at the top. View the html document and compare it the .qmd file before proceeding.

## Getting started

There are two different forms you can edit this document with. "**Source**" uses traditional markdown display and can be helpful for advanced users wanting more flexibility. "**Visual**" is more user friendly so we recommend using it as the editor.

Below is a code chunk. It allows us to insert, display, and run R code in our document. To run a code chunk press the green arrow in the chunk. Try it out!!

```{r}
1 + 3
```

Notice the output is printed below the chunk, with `[1]` signifying a line of output.

```{r}
# Hashtags are used for comments in your code 
# Comments do not get run
2 + 3
4 / 3
```

## Trying out the basics of Markdown:

Try doing the following (you can use either the visual editor to help or the notes provided in lecture earlier today):

- Set up a heading with your name. Under the heading, write a sentence introducing yourself that has a bold word, and a cursive word. Check under the `source` editor how these are defined in the markdown syntax!

## Back to R: variable assignment

```{r}
w <- 0 # Stores the value 0 into the variable w
print(w) # Prints the value stored

x <- "a" # Stores the character/letter "a" into the variable x
x # Also prints the value stored
```

Change the variable "age" to contain your actual age

```{r}
age <- 26
```

Run the following code to print your age to the console.

```{r}
print(paste("My age is ", age))
```

### Committing your changes

At this point we have made a change to this document. You should save your document frequently, and it is up to you how often you commit those changes to Git.

Let's practice committing. Save this document now. In the Git pane commit the change you have made to this document with a meaningful commit message.

After you have committed, "pull" using the blue arrow and then "push" using the green arrow.

Now proceed on through the rest of this lab. Make sure to commit your changes occasionally.

## Vectors

The code below stores the vector containing values 1-5 into variable `y`.

```{r}
y <- c(1, 2, 3, 4, 5)
```

This prints the vector stored in variable y

```{r}
y
```

Computes the length of the vector (i.e. number of elements stored)

```{r}
length(y)
```

Another way to store vectors of consecutive numbers

```{r}
z_vector <- 1:5

z_vector
```

Elements of a vector can be access by specifying its position in the vector.

```{r}
y[1] # prints the first value stored in y
```

Change `new_vector` to contain a vector with the elements: 4, 6, 8, 10.

```{r}
new_vector <- c(4, 6, 7, 10)
```

Print the 3rd (not 1st) value stored in `new_vector`.

## if statements

If statements only run the accompanying code if the condition is true

```{r}
if (w == 0) { 
  print("w equals 0") 
}
```

If else statements provide different code to be run depending on the condition

```{r}
if (w > 0) { 
  print("w is greater than 0") 
} else { 
  print("w is less than 0") 
}
```

Note "!=" means "does not equal"

```{r}
if (age != 999) { 
  print("message 1") 
} else { 
  print("message 2") 
}
```

**Try it out**:

Write an if statement to check if age is less than 21. If true, print the message "I am younger than 21!". If false, print the message "I am 21 or over!". *Make sure to add an R code chunk here using the green +C button above.*

```{r}

```

You can also write if statements with multiple conditions.

```{r}
if (z_vector[2] == 1) {
  print("Stuff")
} else if (z_vector[2] == 2) {
  print("Other stuff")
} else {
  print("More stuff")
}
```

## for loops

For loops iterate over the accompanying code for a given sequence

```{r}
for (iteration in y) { 
  print(iteration) # the value stored in iteration changes with each loop 
}
```

The sequence to iterate over does not need to be defined outside of the loop

```{r}
for (i in 3:5) { 
  print(i) 
}
```

For loops can be helpful for accessing individual elements of a vector

```{r}
for (i in 1:length(new_vector)) { # i will go from 1 to 5 (length of the vector) 
  print(
    paste( 
      "The value stored in position", i, "of new_vector is:", new_vector[i]
    )
  ) 
}
```

`paste()` allows us to print what is stored in a variable. Here we used it within `print()`.

Change the loop below to loop over the days of the week and print the day instead of the month.

```{r}
for (month in c("Jan.", "Feb.", "Mar.", "Apr.", "May", "Jun.", "Jul.", "Aug.", "Sep.", "Oct.", "Nov.", "Dec.")) { 
  
  print(paste("The month is ", month)) 
  
}
```

## Committing again

At this point, we have quite a bit of changes again in the document. If you want to the extra practice with git to build your muscle memory, try committing, and pushing to Github again. Check your Github repo to see if it reflects the changes!!

## Writing functions in R

We can define functions in R in this way:

```{r}
print_name <- function(name){
  print(paste("My name is:",name))
}
print_name("bianca")
```

Try it ! Using `print_name("your_name")` should automatically print your name and let you change it quickly. However, this will not return an object. This means you cannot assign the output of the function to a variable. To do so, we need to use `return` at the end of the function. For example

```{r}
compute_year_born <- function(cur_age, cur_year){
  birth_year <- cur_year - cur_age
  return(birth_year)
}
```

This function takes two arguments, `cur_age`, `cur_year` and return the birth year as an `integer`.

You can use it to assigns to a variable a certain value (similar to what we did above, but using the function to return it instead of plugging it in yourself!)

```{r}
my_birth_year <- compute_year_born(31,2025)
my_birth_year
```

You will note that function does not take into account the fact someone might be born before or after the current date (so that the birth year might be different).

Do you think you can modify the above function to account for that? Hint: you might use an `if` statement as we saw above!! (There is many ways to do this! try out what makes the most sense to you and then compare with your classmates)

## Basic plotting

For a scatterplot we have to specify x and y axis values

```{r}
plot(x = 1:3, y = c(5, 0, 2)) # plots the points (1, 5), (2, 0), (3, 2)
```

We can specify axis labels

```{r}
plot(x = c(-4, 3, 1), y = c(5, 0, 2), xlab = "location", ylab = "height")
```

Make a plot with the following points (1, -3), (-2, 5), (4, 0), (5,1). *You will have to add an R chunk to write your code in!!*

(Add your code here)\

You can change the style of your scatterplot using the argument `pch`. For instance, using my favorite style `pch=16` :

```{r}
plot(x = c(-4, 3, 1), y = c(5, 0, 2), xlab = "location", ylab = "height",pch=16)
```

Play around with changing the number of pch to see how the plot changes!

## Basic plotting : Line plots

Now what if we want to connect the lines in the plot? Use the argument `type="l"`

```{r}
plot(x = c(-4, 1, 3), y = c(5, 0, 2), xlab = "location", ylab = "height",pch=16,type="l")
```

If you want both the points and the line, use the argument `type="b"`

```{r}
plot(x = c(-4, 1, 3), y = c(5, 0, 2), xlab = "location", ylab = "height",pch=16,type="b")
```

Feel free to try different types and see how they look !! (To see what's available, remember to use `?plot` to see what it is possible to do with the function)

## More plotting

If you finish the previous section, type `?plot` to see what is available in terms of options. For instance, try to:

- add a Title to the plot
- try to change the color of the points.
- change the style of the line.

We are going to learn a lot more about plotting tomorrow. These are just some preliminary examples.

\## Wrapping up

We have finished editing this lab now so save, commit, pull, push this document.

Render this document to html and commit, pull, push the html. (Recall, they are two different files!!)
