---
title: "Logistic Regression"
execute:
  echo: true
subtitle: "ISI-BUDS 2026"
format: 
  revealjs:
    slide-number: true
    incremental: true
    theme: ["../slide-style.scss"]
    title-slide-attributes: 
      data-background-size: 12%
      data-background-position: 50% 75%
---

```{r}
#| echo: false
library(tidyverse)
library(bayesrules)
theme_set(theme_gray(base_size = 18))
library(ggplot2)
library(haven)
library(readr)
library(gtsummary)

```




## Introduction

- For linear regression models, the response variable, $Y$, is assumed to be a real-valued continuous random variable. 

- Now consider situations where the response variable is a binary random variable (e.g., disease status).

- For such problems, it is common to use **logistic regression** instead:
$$\begin{eqnarray*}
\log \Big(\frac{\hat{p}}{1- \hat{p}} \Big) & = & a + b_{1}x_1 + \ldots + b_{q}x_{q}
\end{eqnarray*}$$

---

## Logistic regression

- Note that for binary random variables, we have $p = P(Y=1|X)$; that is, $p$ is the probability of the outcome of interest (denoted as 1) given the explanatory variables.

- The term $\Big(\frac{\hat{p}}{1- \hat{p}} \Big)$ is called the **odds** of $Y=1$. 

- The term $\log \Big(\frac{\hat{p}}{1- \hat{p}} \Big)$, i.e.,  log of odds, is called the **logit** function. 

- Although $p$ is a real number between 0 and 1, its logit transformation can be any real number between $-\infty$ to $+\infty$.



---

## Logistic regression

- We can exponentiate both sides,
$$\begin{eqnarray*}
\frac{\hat{p}}{1- \hat{p}}  & = & \exp(a + b_{1}x_1 + \ldots + b_{q}x_{q})
\end{eqnarray*}$$

- Then, we can find $\hat{p}$ using the **logistic function**:
$$\begin{eqnarray*}
\hat{p} & = & \frac{\exp(a + b_{1}x_1 + \ldots + b_{q}x_{q})}{1 + \exp(a + b_{1}x_1 + \ldots + b_{q}x_{q})}
\end{eqnarray*}$$





## Smoking and Lowbirth Babies


- As an example, we use the _birthwt_ data set to model the relationship between having low birthweight babies (a binary variable), $Y$, and smoking during pregnancy, $X$. 

- The binary variable _low_ identifies low birthweight babies (_low_ = 1 for low birthweight babies, and 0 otherwise). 

- The binary variable _smoke_ identifies mothers who were smoking during pregnancy (_smoke_ =1 for smoking during pregnancy, and 0 otherwise).


## Logistic Regression Models Using R

```{r}
library(MASS)
data("birthwt")
model <- glm(low ~ smoke, data = birthwt, family = binomial)
summary(model)
```







## Smoking and Lowbirth Babies


- For the above example, the estimated values of the intercept $\alpha$ and the regression coefficient $\beta$ are $a=-1.09$ and $b=0.70$ respectively. 

- Therefore, 
\begin{eqnarray*}
\frac{\hat{p}_{x}}{1- \hat{p}_{x}} & = & \exp(-1.09 + 0.70x)
\end{eqnarray*}

- Here, $\hat{p}_{x}$ is the estimated probability of having a low birthweight baby for a given $x$. 

- The left-hand side of the above equation is the estimated **odds** of having a low birthweight baby. 


## Smoking and Lowbirth Babies


- For non-smoking mother, $x=0$, the odds of having low birthweight baby is 
\begin{eqnarray*}
\frac{\hat{p}_{0}}{1- \hat{p}_{0}} & = & \exp(-1.09) \\
& = & 0.34
\end{eqnarray*}

- That is, the exponential of the intercept is the odds when $x=0$, which is sometimes referred to as the **baseline odds**. 



## Smoking and Lowbirth Babies

- For mothers who smoke during pregnancy, $x=1$, and
\begin{eqnarray*}
\frac{\hat{p}_{1}}{1- \hat{p}_{1}} & = &  \exp(-1.09 + 0.7)\\
& = &  \exp(-1.09) \exp(0.7)
\end{eqnarray*}


- As we can see, corresponding to one unit increase in $x$ from $x=0$ (non-smoking) to $x=1$ (smoking), the odds multiplicatively increases by the exponential of the regression coefficient. 


## Smoking and Lowbirth Babies

- Note that
\begin{eqnarray*}
\frac{\frac{\hat{p}_{1}}{1- \hat{p}_{1}}}{\frac{\hat{p}_{0}}{1- \hat{p}_{0}}} & = &  \frac{\exp(-1.09) \exp(0.7)}{\exp(-1.09)}\\
& = & \exp(0.7)
\end{eqnarray*}

- We can interpret the exponential of the regression coefficient as the **odds ratio** of having low birthweight babies for smoking mothers compared to non-smoking mothers. 

- Here, the estimated odds ratio is $\exp(0.7) = 2.01$ so the odds of having a low birthweight baby almost doubles for smoking mothers compared to non-smoking mothers. 



## Odds Ratio


- In general, 
  - if $\beta>0$, then $\exp(\beta) > 1$ so the odds increases as $X$ increases;
  - if $\beta<0$, then $0 < \exp(\beta) < 1$ so the odds decreases as $X$ increases;
  - if $\beta=0$, the odds ratio is 1 so the odds does not change with $X$ according to the assumed model. 





## Confidence Interval

- Given the point estimates $a$ and $b$, we find the confidence intervals for the regression parameters as follows:
\begin{eqnarray*}
&[a - z_{\mathrm{crit}} \times \mathit{SE}_{a}, a + z_{\mathrm{crit}} \times \mathit{SE}_{a}]&\\
&[b - z_{\mathrm{crit}} \times \mathit{SE}_{b}, b + z_{\mathrm{crit}} \times \mathit{SE}_{b}]&
\end{eqnarray*}

- Here, $\mathit{SE}_{a}$ and $\mathit{SE}_{b}$ are the standard errors of our estimators, and $z_{\mathrm{crit}}$ is the $z$-critical value for the required confidence level. 


## Confidence Interval

- For our example, the 95\% confidence intervals of $\alpha$ is
\begin{eqnarray*}
[-1.09 - 2 \times 0.21, -1.09 + 2 \times 0.21] & = & [-1.51, -0.67]. 
\end{eqnarray*}

- Because the baseline odds (i.e., the odds when $x=0$) is $\exp(\alpha)$, the 95\% confidence interval for the baseline odds is obtained as follows:
\begin{eqnarray*}
[\exp(-1.51), \exp(-0.67)]  & = & [0.22, 0.51]. 
\end{eqnarray*}

- The 95\% confidence interval of $\beta$ is
\begin{eqnarray*}
[0.70 - 2 \times 0.32, 0.70 +  2 \times 0.32] & = & [0.06, 1.34].
\end{eqnarray*}

## Confidence Interval

- Because the odds ratio is $\exp(\beta)$, the 95\% confidence interval of the odds ratio is obtained as follows:
\begin{eqnarray*}
[\exp(0.06), \exp(1.34)]& = & [1.06, 3.82].
\end{eqnarray*}



## Confidence Interval Using R
We can use R to obtain the confidence intervals for the regression parameters, baseline odds, and odds ratio.
```{r}
confint(model, level = 0.95)
exp(confint(model, level = 0.95))
```

## Hypothesis Testing

- The regression coefficient $\beta$ captures the relationship between the response variable $Y$ and the explanatory variable $X$. 

- In logistic regression model, the assumption is that if the two variables are related, log odds for the response variable changes linearly with $X$. 

- Using logistic regression models, we can specify our hypothesis regarding the relationship between $X$ and $Y$ in terms of $\beta$ as $H_{A}: \beta >0$, or $H_{A}: \beta < 0$, or $H_{A}: \beta \ne 0$; It is common to use two-sided alternative hypotheses. 

- In contrast the null hypothesis is specified as $H_{0}: \beta = 0$. 


## Hypothesis Testing

- As before, we use the observed significance level, i.e., $p$-value, to decide whether we should reject the null hypothesis. 

- To do this, we first need to find the $z$-score by dividing $b$ (i.e., the point estimate of $\beta$) by its standard error. 

- Then, we find the $p$-value by
calculating the probability of as or more extreme values than the $z$-score under the null.

## Hypothesis Testing

- To assess the null hypothesis $H_{0}: \beta = 0$, we first calculate the $z = b/\mathit{SE}_{b}$
and find the corresponding $p$-value as follows:
\begin{array}{l@{\quad}l}
\mbox{if}\ H_{A}: \beta < 0, & p_{\mathrm{obs}} = P(Z \leq z), \\
\mbox{if}\ H_{A}: \beta > 0, & p_{\mathrm{obs}} = P(Z \geq z ), \\
\mbox{if}\ H_{A}: \beta \ne 0, & p_{\mathrm{obs}} = 2 \times P\bigl(Z \geq | z | \bigr),
\end{array}
where $Z$ has the standard normal distribution.



## Hypothesis Testing

- For the above example
\begin{equation*}
z  =  \frac{0.70}{0.32} = 2.2.
\end{equation*}
This is of course what R provides under the ``z value'' column. 

- The corresponding p-value is 0.028, which is also provided by R under ``Pr(>|z|)''.

- In this example, we can reject the null hypothesis at 0.05 significance level. 


## Prediction

- We can use logistic regression models for predicting the unknown values of the response variable $Y$ given the value of the predictor value $X$.
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(a + bx)}{1 + \exp(a + bx)}
\end{eqnarray*}

- For the above example,
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(-1.09 + 0.70x)}{1 + \exp(-1.09 + 0.70x)}
\end{eqnarray*}


## Prediction

- Therefore, the estimated probability of having a low birthweight baby for non-smoking mothers, $x=0$, is
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(-1.09)}{1 + \exp(-1.09)} = 0.25
\end{eqnarray*}

- This probability increases for mothers who smoke during pregnancy,
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(-1.09 + 0.7)}{1 + \exp(-1.09 + 0.7)} = 0.40
\end{eqnarray*}

- That is, the risk of having a low birthweight baby increases by 60\% if a mother smokes during her pregnancy. 





## Logistic Regression with One Numerical Predictor

- For the most part, we follow similar steps to fit the model, estimate regression parameters, perform hypothesis testing, and predict unknown values of the response variable. 

- As an example, we want to investigate the relationship between having a low birthweight baby, $Y$, and mother's age at the time of pregnancy, $X$. 

## Mother's Age and Lowbirth Babies

```{r}
model2 <- glm(low ~ age, data = birthwt, family = binomial)
summary(model2)
```




## Mother's Age and Lowbirth Babies

- To interpret $b$, consider mothers whose age is 20 years old at the time of pregnancy, 
\begin{eqnarray*}
\log \Big(\frac{\hat{p}_{20}}{1- \hat{p}_{20}} \Big) & = & 0.38 - 0.05 \times 20\\ 
\frac{\hat{p}_{20}}{1- \hat{p}_{20}} & = &  \exp(0.38 - 0.05 \times 20)\\
& = &  \exp(0.38) \exp(- 0.05 \times 20)
\end{eqnarray*}

## Mother's Age and Lowbirth Babies

- For mothers who are one year older (i.e., one unit increase in age):
\begin{eqnarray*}
\log \Big(\frac{\hat{p}_{21}}{1- \hat{p}_{21}} \Big) & = & 0.38 - 0.05 \times 21\\ 
\frac{\hat{p}_{21}}{1- \hat{p}_{21}} & = &  \exp(0.38 - 0.05 \times 21)\\
& = &  \exp(0.38) \exp(- 0.05 \times 21)
\end{eqnarray*}


## Mother's Age and Lowbirth Babies


- The odds ratio for comparing 21 year old mothers to 20 year old mothers is
\begin{eqnarray*}
\frac{\frac{\hat{p}_{21}}{1- \hat{p}_{21}}}{\frac{\hat{p}_{20}}{1- \hat{p}_{20}}} & = &  \frac{ \exp(0.38) \exp(- 0.05 \times 21))}{ \exp(0.38) \exp(- 0.05 \times 20)}\\
& = & \exp(- 0.05 \times 21 + 0.05 \times 20)\\
& = & \exp(- 0.05)
\end{eqnarray*}

- Therefore, $\exp(b)$ is the estimated odds ratio comparing 21 year old mothers to 20 year old mothers. 

## Logistic Regression with One Numerical Predictor

- In general, $\exp(b)$ is the estimated odds ratio for comparing two subpopulations, whose predictor values are $x+1$ and $x$,
\begin{eqnarray*}
\frac{\frac{\hat{p}_{x+1}}{1- \hat{p}_{x+1}}}{\frac{\hat{p}_{x}}{1- \hat{p}_{x}}} & = & \exp(b)
\end{eqnarray*}



## Mother's Age and Lowbirth Babies

- As before, we can use the estimated regression parameters to find $\hat{p}_{x}$ and predict the unknown value of the response variable. 
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(a + bx)}{1 + \exp(a + bx)}\\
& = & \frac{\exp(0.38 -0.05x)}{1 + \exp(0.38 -0.05x)}.
\end{eqnarray*}

## Mother's Age and Lowbirth Babies

- For example, for mother who are 20 years old at the time of pregnancy, the estimated probability of having a low birthweight baby is
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(0.38 -0.05 \times 20)}{1 + \exp(0.38 -0.05 \times 20)} = 0.35.
\end{eqnarray*}


- Caution: Depending on the sample selection strategy, interpreting this probability may become problematic.



# **Multiple Logistic Regression**



## Multiple Logistic Regression


- We now discuss logistic regression models with multiple explanatory (predictor) variables. 

- For example, suppose we want to investigate the relationship between smoking during pregnancy and having a low birthweight baby, but we suspect that mother's age at the time of pregnancy might influence this relationship. 
  \begin{eqnarray*}
  \log \Big(\frac{p}{1- p} \Big) & = & \alpha + \beta_{1} \textrm{smoke} + \beta_{2} \textrm{age} 
  \end{eqnarray*}
  where $p$ is the probability of having a low birthweight baby for given values of _smoke_ and _age_.
  
  



## Multiple Logistic Regression
```{r}
model3 <- glm(low ~ smoke + age, data = birthwt, family = binomial)
summary(model3)
```



## Multiple Logistic Regression
  
  
  - As before, finding confidence intervals and performing hypothesis testing remain similar to what we discussed for models with a single predictor, so we focus on prediction and interpretation of point estimates. 
  
  - For the above example, the estimated value of $\alpha$ is $a=0.06$. 
  
  - In general, this could be interpreted as the odds when all predictors are set to zero. In this case, however, mother's age cannot be zero.



## Multiple Logistic Regression



- The estimated value of $\beta_{1}$, i.e., the regression coefficient of smoke, is $b_{1} = 0.69$. 

- In general, this is interpreted in terms of the odds ratio corresponding to one unit increase in its related predictor \emph{when all other predictors are kept constant}. 

- Here, $\exp(0.69) = 1.99$ is the estimated odds ratio of having a low birthweight baby comparing smoking to non-smoking mothers with the same age (i.e., keeping age constant). 

- Therefore, the odds of having a low birthweight baby almost doubles for smoking mothers compared to non-smoking mothers of the same age. 




## Multiple Logistic Regression

- The estimated value of $\beta_{2}$, i.e., the regression coefficient of age, is $b_{2} = -0.05$. 

- The estimated odds ratio for comparing two groups of mothers with 1 year age difference (e.g., comparing 20 year old mothers to 19 year old mothers) with the same smoking status (i.e., keeping _smoke_ constant) is $\exp(-0.05) = 0.95$. 

- Of course, this result is not statistically signifiant since $p$-value=0.12 is above the commonly used significance levels. 



## Multiple Logistic Regression

- We can use our model to estimate the risk of having a low birthweight baby,
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(a + b_1 \mathrm{smoke} + b_{2} \mathrm{age})}{1 + \exp(a + b_1 \mathrm{smoke} + b_{2} \mathrm{age})}
\end{eqnarray*}

- For a 20 year old mother who smokes during her pregnancy, the estimated probability of having a low birthweight baby is
\begin{eqnarray*}
\hat{p}_{x} & = & \frac{\exp(0.06 + 0.69 \times 1 -0.05 \times 20)}{1 + \exp(0.06 + 0.69 \times 1  -0.05 \times 20)}\\
& = & 0.44
\end{eqnarray*}



::: {.v-center}

# **Model Assessment and Predictive Modeling**

:::


## Model Assessment

- We typically build regression models either for hypothesis testing or prediction. 

- When our objective is to perform hypothesis testing, the set of variables we include in our model is dictated by our hypothesis and our domain knowledge. 




## Goodness-of-fit


- The **deviance** is a common measure of discrepancy (i.e., lack of fit) between the model and the observed data: the lower the deviance, the better the fit.

- The value of deviance is provided by R .

- The fit (and deviance) improves as we increase the number of predictors. 

- Akaike proposed to penalize against model complexity by adding $2k$, where $k$ is the number of model parameters, to the deviance. 

- The resulting measure is called Akaike Information Criterion (AIC),
\begin{eqnarray*}
AIC & = &  Deviance  + 2k
\end{eqnarray*}



## Predictive Modeling

- When our objective is to predict the unknown values of the response variable, we should include in the systematic component any predictor that would improve the predictive power of our model.

- A common measure for predictive power is _accuracy rate_, which is defined as the proportion of the
times the correct category (0 or 1 in this case) is predicted for future observations (or observations in the test set).




## Predictive Modeling

- Note that the outputs of logistic regression models are in fact between 0 and 1, which are interpreted as probabilities. 

- Therefore, we need to set an appropriate cutoff to obtain binary predictions, $\hat{y}$.

- In general, the cutoff depends on the loss function; that is, the cost of predicting the class as 0, when the true class is 1, and vice versa. 

- For simplicity, it is common to assign a test case to the class with the highest probability; that is, we set the cutoff at $0.5$.


## Predictive Modeling


- Instead of averaging over all predictions, it might be more informative to separate the types of error. 

- One common approach for doing this is to present the results in the form of a _confusion matrix_.

## Predictive Modeling


:::: {.table caption="Confusion Matrix"}

+--------------+-----+----------------+----------------+
|              |     | Predicted: 0   | Predicted: 1   |
+==============+=====+================+================+
| True Class   | 0   | True Negative  | False Positive |
+--------------+-----+----------------+----------------+
|              | 1   | False Negative | True Positive  |
+--------------+-----+----------------+----------------+

::::


- True positive rate is also called **sensitivity**

- True negative rate is called **specificity**.


## Predictive Modeling



- Instead of setting an arbitrary cutoff (e.g., 0.5), we can use Receiver Operating Characteristic (ROC) curves, which allow for simultaneous consideration of sensitivity and specificity without setting an arbitrary cut-off. 

- The curve plots true positive rate (sensitivity) as a function of false positive rate (1-specificity) for all possible cutoffs from 0 to 1. 



## Predictive Modeling


- The area under the curve (AUC) is a measure of overall performance: higher AUC corresponds to higher predictive power.

- If the AUC is one, then the value of the response variable is always predicted correctly. 

- If it is 0.5, the predictive model is equivalent to tossing a fair coin, e.g. useless.

- R has several packages (e.g., pROC, ROCR) for plotting ROC curves and calculating their AUC.


## Predictive Modeling

```{r}
#| echo: false
library(pROC)

birthwt$p <- predict(model3, type = "response")

roc_obj <- roc(birthwt$low, birthwt$p)

roc_df <- data.frame(
  FPR = 1 - roc_obj$specificities,
  TPR = roc_obj$sensitivities
)


ggplot(roc_df, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", size = 1.2) +
  geom_abline(linetype = "dashed", color = "gray") +
  labs(
    title = paste("ROC Curve (AUC =", round(auc(roc_obj), 3), ")"),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal()

```


