---
title: "Day 11 Lab: Logistic Regression Solutions"
format: 
  html:
    toc: true
    toc-expand: true
    toc-depth: 2
embed-resources: true
editor: visual
execute:
  echo: fenced
theme: cerulean
---

Suzuki et al. (2006) measured sand grain size on 28 beaches in Japan and observed the presence or absence of the burrowing wolf spider *Lycosa ishikariana* on each beach.
One goal of this study was to determine whether there was a relationship between sand grain size (mm) and the presence or absence of the species, in hopes of understanding more about the biology of the spiders.
If there is a relationship between the two variables, it would be sand grain size affecting spiders, not the presence of spiders affecting the sand.
Because this species is endangered, another goal would be to find an equation that would predict the probability of a wolf spider population surviving on a beach with a particular sand grain size, to help determine which beaches to reintroduce the spider to[^1].

[^1]: [McDonald, J.H. 2014. Handbook of Biological Statistics (3rd ed.). Sparky House Publishing, Baltimore, Maryland.](https://www.biostathandbook.com/simplelogistic.html)

```{r}
library(tidyverse)
set.seed(630)
```

## Question 1

Write out an appropriate regression equation to investigate the association between spider presence and sand grain size.

## Question 1 Solution

$$\log(\frac{p_i}{1 - p_i}) = \beta_0 + \beta_1 \text{Grain Size}_i,$$ where $p_i = E[\text{Spider Presence}_i | \text{Grain Size}_i]$.

## Question 2

Read in the spider data.
We need the response variable to be coded as 0's and 1's.
Run the chunk below to make a new variable recording spider presence.

```{r}
spider_data <- read_csv("data/spider-data.csv")
spider_data$present <- ifelse(spider_data$spiders == "present", yes = 1, no = 0)
```

Fit the regression equation you specified in question 1 and write out the estimated regression equation.

## Question 2 Solution

```{r}
spider_model <- glm(
  present ~ grain_size,
  data = spider_data,
  family = binomial
)

(spider_model_sum <- summary(spider_model))
```

$$\log(\frac{\hat{p}_i}{1 - \hat{p}_i}) = -1.65 + 5.12 * \text{Grain Size}_i,$$

## Question 3

Solve your estimate regression equation for $\hat{p}_i$.

## Question 3 Solution

$$\log(\frac{\hat{p}_i}{1 - \hat{p}_i}) = -1.65 + 5.12 * \text{Grain Size}_i$$ $$\frac{\hat{p}_i}{1 - \hat{p}_i} = exp(-1.65 + 5.12 * \text{Grain Size}_i)$$ $$\hat{p}_i = exp(-1.65 + 5.12 * \text{Grain Size}_i) - \hat{p}_i \times exp(-1.65 + 5.12 * \text{Grain Size}_i)$$ $$\hat{p}_i(1 + exp(-1.65 + 5.12 * \text{Grain Size}_i)) = exp(-1.65 + 5.12 * \text{Grain Size}_i)$$ $$\hat{p}_i = \frac{exp(-1.65 + 5.12 * \text{Grain Size}_i)}{(1 + exp(-1.65 + 5.12 * \text{Grain Size}_i))}.$$

## Question 4

Compute 95% confidence intervals for both coefficients and interpret your estimates of $\beta_0$ and $\beta_1$ in context.

## Question 4 Solution

```{r}
beta0 <- spider_model_sum$coefficients["(Intercept)", "Estimate"]
beta0_se <-  spider_model_sum$coefficients["(Intercept)", "Std. Error"]

beta1 <- spider_model_sum$coefficients["grain_size", "Estimate"]
beta1_se <- spider_model_sum$coefficients["grain_size", "Std. Error"]

beta0_95ci <- beta0 + c(-1, 1) * qnorm(1 - 0.05 / 2) * beta0_se
beta1_95ci <- beta1 + c(-1, 1) * qnorm(1 - 0.05 / 2) * beta1_se
```

We estimate the log odds of spider presence for beaches with sand grain size of zero to be `r round(beta0, 2)` (95% CI (`r round(beta0_95ci[1], 2)`, `r round(beta0_95ci[2], 2)`)).

We estimate a difference in log odds of spider presence of `r round(beta1, 2)` (95% CI (`r round(beta1_95ci[1], 2)`, `r round(beta1_95ci[2], 2)`)), associated with a one millimeter increase in sand grain size.

## Question 5

Compute 95% confidence intervals for both $exp(\beta_0)$ and $exp(\beta_1)$ and interpret your estimates in context.

## Question 5 Solution

```{r}
exp_beta0 <- exp(beta0)
exp_beta0_95ci <- exp(beta0_95ci)

exp_beta1 <- exp(beta1)
exp_beta1_95ci <- exp(beta1_95ci)
```

We estimate the odds of spider presence for beaches with sand grain size of zero to be `r round(exp_beta0, 2)` (95% CI (`r round(exp_beta0_95ci[1], 2)`, `r round(exp_beta0_95ci[2], 2)`)).

We estimate an odds ratio of spider presence of `r round(exp_beta1, 2)` (95% CI (`r round(exp_beta1_95ci[1], 2)`, `r round(exp_beta1_95ci[2], 2)`)), associated with a one millimeter increase in sand grain size.

# Wald vs LRT test 

Checking how to compute these confidence intervals in R...

```{r}
# LRT test
confint(spider_model)
confint(spider_model,test = "LRT")
```

```{r}
# Wald test
confint.default(spider_model)
```
