---
title: "Day 7 Lab Monte Carlo Methods in Inference"
format: 
  html:
    toc: true
    toc-expand: true
    toc-depth: 2
editor: visual
embed-resources: true
execute:
  echo: fenced
theme: cerulean
---

# Mean Squared Error (MSE) and Bias

```{r}
set.seed(5172013)
```

Simulate 1000 data sets with 30 $Poisson(\lambda = 2.3)$ realizations in each data set

```{r}
sim_size <- 1000
sample_size <- 30
pois_lambda <- 2.3

pois_data <- matrix(
  rpois(n = sim_size * sample_size, lambda = pois_lambda),
  nrow = sim_size, 
  ncol = sample_size
)
```

Define two types of estimators:

```{r}
pois_est1 <- function(pois_counts){
  return(mean(pois_counts))
}

pois_est2 <- function(pois_counts){
  return(var(pois_counts))
}
```

Apply the two estimators to the simulated data

```{r}
est_vec1 <- numeric(sim_size)
est_vec2 <- numeric(sim_size)

for (i in 1:sim_size){
  est_vec1[i] <- pois_est1(pois_data[i,])
  est_vec2[i] <- pois_est2(pois_data[i,])
}
```

Plot box plots of the two sampling distributions

```{r}
library(ggplot2)

est_df <- data.frame(
  Estimator = c(rep("Est 1", length(est_vec1)), rep("Est 2", length(est_vec2))),
  Estimate = c(est_vec1, est_vec2)
)

ggplot(data = est_df, mapping = aes(x = Estimator, y = Estimate)) +
  geom_boxplot() +
  geom_hline(yintercept = pois_lambda, color = "darkgreen", linetype = "dashed", size = 1.5) +
  theme_bw()
```

Compute MSE for both estimators

```{r}
(mse1 <- mean((est_vec1 - pois_lambda)^2))
(mse2 <- mean((est_vec2 - pois_lambda)^2))
```

Compute Monte Carlo errors

```{r}
(mse1_mc_error <- sd((est_vec1 - pois_lambda)^2) / sqrt(sim_size))
(mse2_mc_error <- sd((est_vec2 - pois_lambda)^2) / sqrt(sim_size))
```

and intervals for the error

```{r}
(c(mse1 - 1.96 * mse1_mc_error, mse1 + 1.96 * mse1_mc_error))
(c(mse2 - 1.96 * mse2_mc_error, mse2 + 1.96 * mse2_mc_error))
```

What do we notice?

## Exercise 1

a.  Compute biases of the above estimators via Monte Carlo.
    Can you say with some confidence whether these biases are 0 or not?

b.  Compute coverage of the 95% confidence intervals of the first estimator (the one that uses sample mean).
    Can you tell if your estimated coverage is different from 0.95?

## Exercise 1 Solutions

a.  

Compute bias for both estimators

```{r}
(bias1 <- mean(est_vec1 - pois_lambda))
(bias2 <- mean(est_vec2 - pois_lambda))
```

Compute Monte Carlo errors

```{r}
(bias1_mc_error <- sd(est_vec1-pois_lambda) / sqrt(sim_size))
(bias2_mc_error <- sd(est_vec2-pois_lambda) / sqrt(sim_size))
```

and 95% confidence intervals

```{r}
c(bias1 - 1.96 * bias1_mc_error, bias1 + 1.96 * bias1_mc_error)
c(bias2 - 1.96 * bias2_mc_error, bias2 + 1.96 * bias2_mc_error)
```

b.  

Some guidelines:

Write a function that returns 95% confidence intervals for the first estimator

```{r}
pois_ci <- function(pois_counts){
  sample_mean <- mean(pois_counts)
  sample_sd <- sd(pois_counts)
  
  c(
    "lb" = sample_mean - 1.96 * sample_sd / sqrt(length(pois_counts)),
    "ub" = sample_mean + 1.96 * sample_sd / sqrt(length(pois_counts))
  )
}
```

Allocate a vector for 0/1's to indicate coverage of the true value

```{r}
coverage_indicator <- numeric(sim_size)

for (i in 1:sim_size){
  cur_ci <- pois_ci(pois_data[i,])
  
  if ((cur_ci[1] < pois_lambda) && (cur_ci[2] > pois_lambda)){
    coverage_indicator[i] <- 1
  }
}

(est_cov <- mean(coverage_indicator))
```

Monte Carlo error -- notice the placement of the square root!

```{r}
est_cov_error <- sqrt(est_cov * (1 - est_cov) / sim_size)

c(est_cov - 1.96 * est_cov_error, est_cov + 1.96 * est_cov_error)
```

# Hypothesis Testing and Confidence Intervals

## Example 1: Hypothesis testing

Suppose we are interested in making inference about the mean of a normally distributed variable with known standard deviation of 100.
For our inference, $H_0: \mu = 500$; $H_a: \mu > 500$.
We wish to calculate the empirical probability of making a type I error.

```{r}
set.seed(378923)

samp_size <- 20
alpha <- 0.05
mu0 <- 500
sigma <- 100

sim_size <- 1000           # number of replicates
p <- numeric(sim_size)      # storage for p-values

for (sim in 1:sim_size) {
  samp <- rnorm(samp_size, mu0, sigma) # simulate from H0
  ttest <- t.test(samp, alternative = "greater", mu = mu0)
  p[sim] <- ttest$p.value
}

(p_hat <- mean(p < alpha))
(se_hat <- sqrt(p_hat * (1 - p_hat) / sim_size))
p_hat + c(-1, 1) * 1.96 * se_hat # c(-1, 1) vectorizes plus and minus
```

Since the data are normal to begin with, we would expect 0.05 to be contained in the interval

## Example 2: Power

Now we will consider the same case as example 1, but we are interested in the probability of making a type II error.

Recall: a type II error ($\beta$) occurs in hypothesis testing when a **false null hypothesis** is **not** rejected.

This means failing to detect a real effect or difference that actually exists, i.e. when the true mean is different from $H_0$.

Power = 1-$\beta$.

```{r}
mu <- seq(from = 450, to = 650, by = 10)  #alternatives
n_mu <- length(mu)
power <- numeric(n_mu)

for (i in 1:n_mu) {
  mu_a <- mu[i] # Select an alternative
  p <- numeric(sim_size)
  
  for(sim in 1:sim_size) {
    # simulated from alternative
    samp <- rnorm(samp_size, mu_a, sigma) 
    # perform t-test
    ttest <- t.test(samp, mu = mu0, alternative = 'greater') #change to two.sided
    p[sim] <- ttest$p.value
  }
  
  power[i] <- mean(p < 0.05) # We want low p-values since the null is not true.
  
}
```

```{r}
se <- sqrt(power * (1 - power) / sim_size)

power_df <- data.frame(
  "mu" = mu, 
  "power" = power,
  "ci_lower_bound" = power - 1.96 * se,
  "ci_upper_bound" = power + 1.96 * se
)

ggplot(data = power_df, aes(x = mu, y = power)) +
  geom_point(color = "red") +
  geom_line(linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  geom_vline(xintercept = mu0, linetype = "dotted") +
  geom_errorbar(aes(x = mu, ymin = ci_lower_bound, ymax = ci_upper_bound)) +
  labs(x = expression(mu)) +
  theme_bw() +
  annotate(x = 500, y = 0.65, geom = "text", label = "H[0]", parse = TRUE) +
  annotate(x = 595, y = 0.05, geom = "text", label = "alpha", parse = TRUE) 
```

## Exercise 1

Suppose the heights of Douglas Firs in a stand of trees planted the same year are normally distributed with a height of 20 feet and a standard deviation of 4 feet.
However, you have reason to believe that the soil quality is higher in an area near a stream, so you think the trees might grow faster there.
You have the time and money to sample the heights of 15 trees near the stream.
Following convention, you plan on using an $\alpha$ of 0.05.

a.  What is the empirical probability of a type I error, including a 95% confidence interval for that probability?

b.  Calculate the power of the test for a range of reasonable alternatives.

## Exercise 1 Solution

a.  

```{r}
samp_size <- 15
alpha <- 0.05
mu0 <- 20
sig <- 4

p <- numeric(sim_size)

for(sim in 1:sim_size) {
  samp <- rnorm(samp_size, mean = mu0, sd = sig)
  ttest <- t.test(samp, alternative = 'greater', mu = mu0)
  p[sim] <- ttest$p.value
}

(p_hat <- mean(p < alpha))
se_hat <- sqrt(p_hat * (1 - p_hat) / sim_size)
p_hat + 1.96 * c(-1, 1) * se_hat
```

b.  

```{r}
mu <- seq(from = 18, to = 25, by = 0.5)  #alternatives
n_mu <- length(mu)
power <- numeric(n_mu)

for (i in 1:n_mu) {
  mu_a <- mu[i] # Select an alternative
  p <- numeric(sim_size)
  
  for(sim in 1:sim_size) { 
    samp <- rnorm(samp_size, mean = mu_a, sd = sig)
    ttest <- t.test(samp, mu = mu0, alternative = 'greater') 
    p[sim] <- ttest$p.value
  }
  
  power[i] <- mean(p < 0.05)
  
}
```

```{r}
se <- sqrt(power * (1 - power) / sim_size)

power_df <- data.frame(
  "mu" = mu, 
  "power" = power,
  "ci_lower_bound" = power - 1.96 * se,
  "ci_upper_bound" = power + 1.96 * se
)

ggplot(data = power_df, aes(x = mu, y = power)) +
  geom_point(color = "red") +
  geom_line(linetype = "dashed") +
  geom_hline(yintercept = 0.05, linetype = "dotted") +
  geom_vline(xintercept = mu0, linetype = "dotted") +
  geom_errorbar(aes(x = mu, ymin = ci_lower_bound, ymax = ci_upper_bound)) +
  labs(x = expression(mu)) +
  theme_bw() +
  annotate(x = mu0, y = 0.65, geom = "text", label = "H[0]", parse = TRUE) +
  annotate(x = 22.5, y = 0.05, geom = "text", label = "alpha", parse = TRUE)
```

# Bootstrap

## Example

A company is producing a medical patch that infuses a hormone into the blood.
The company would like to open a new manufacturing plant, but the Food and Drug Administration (FDA) requires that the company establishes bioequivalence of patches produced at the old and new plants.
To show the bioequivalence, the company recruits n subjects, each of whom at different times will wear three different patches:

1.  a placebo patch that infuses no hormone
2.  an "old" patch produced at the old plant
3.  an "new" patch produced at the new plant

The company needs to demonstrate that on average the different between the amounts of hormone infused by "new" and "old" patch is small compare to the difference in this quantity between the "old" and placebo patches.
More precisely, the company needs to show that $$\frac{|E(new) − E(old)|}{|E(old) − E(placebo)|} < 0.2 $$

load the patch data from the class web site

```{r}
patch_data <- readr::read_table("data/patch_data.txt", show_col_types = FALSE)
(patch_data)
```

Let's define a function that computes the ratio of mean differences

```{r}
ratio_stat <- function(data, indices){
  return(
    mean(data$newpatch[indices] - data$oldpatch[indices]) / mean(data$oldpatch[indices] - data$placebo[indices])
  ) 
}

subj_num <- dim(patch_data)[1]

obs_est <- ratio_stat(patch_data, c(1:subj_num))

patch_boot_size <- 10000
```

prepare a vector for mean ratio estimates

```{r}
patch_boot_est <- numeric(patch_boot_size)

for (i in 1:patch_boot_size){
  ## sample with replacement rows of the patch_data matrix
  new_rows <- sample(1:subj_num, size = subj_num, replace = TRUE)
  
    ## re-estimate the ratio of mean differences
  patch_boot_est[i] <- ratio_stat(patch_data, new_rows)
}

patch_boot_est_df <- data.frame("patch_boot_est" = patch_boot_est)

ggplot(data = patch_boot_est_df, aes(x = patch_boot_est)) +
  geom_histogram(fill = "gray80", color = "black") +
  geom_vline(xintercept = obs_est, color = "red", linewidth = 1.5) +
  labs(x = "Patch Ratio", "Count") +
  theme_bw()
  
```

Let's calculate the estimator's standard error

```{r}
boot_patch_sd <- sd(patch_boot_est)
```

We can calculate the bias:

```{r}
mean(patch_boot_est - obs_est)
```

Let's form a basic bootstrap confidence interval

```{r}
(norm_int <- obs_est + 1.96 * c(-1, 1) * sd(patch_boot_est))
```

Let's form a quantile bootstrap confidence interval

```{r}
(quant_int <- quantile(patch_boot_est, probs = c(0.025, 0.975)))
```

Stop to visit help file for the "quantile" function
