---
title: "Day 7 Lab: Monte Carlo Integration"
format: 
  html:
    toc: true
    toc-expand: true
    toc-depth: 2
editor: visual
execute:
  echo: fenced
embed-resources: true
---

# Monte Carlo integration

```{r}
set.seed(5172013)
```

## Example 1: integral of $e^{x^4 - x^2}$ over \[0,1\]

Define an integrand

```{r}
integrand_function <- function(x){
  return(exp(x^4 - x^2))
}
```

Plot the integrand on the interval \[0,1\]

```{r}
x_values <- c(0:100) / 100

plot(x_values, integrand_function(x_values), type = "l", lwd = 2, xlab = "x", ylab = "y", cex.lab = 1.3, col = "red")
text(0.6, 0.95, expression(y == e^{x^4 - x^2}), cex = 2)
```

See `?plotmath` for more info on plotting math in R

### Deterministic numerical integration from 0 to 1

Notice that we are passing function `integrand_function` as an argument Also the integral can be seen as $E(e^{U^4 - U^2})$, where $U \sim Uniform(0, 1)$

```{r}
det_int <- integrate(integrand_function, lower = 0, upper = 1)
```

The function `integrate()` returns an object, not a value to find out what is in the object, type `names(det_int)`

```{r}
names(det_int)

det_int$value
det_int$message
```

### Monte Carlo integration

By default `runif()` defines uniform distribution on \[0,1\]

```{r}
mcarlo_iterations <- 1000000

unif_sample <- runif(mcarlo_iterations)
```

artihmetic average

```{r}
(mcarlo_int = mean(integrand_function(unif_sample)))
```

Monte Carlo error

```{r}
(mcarlo_error = sd(integrand_function(unif_sample))/sqrt(mcarlo_iterations))
```

Monte Carlo 95% confidence interval

```{r}
(c(mcarlo_int - 1.96*mcarlo_error, mcarlo_int + 1.96*mcarlo_error))
```

Compare to the deterministic numerical integration

```{r}
(det_int)
```

## Example 2: second moment of the Beta distribution $X \sim Beta(\alpha, \beta)$

```{r}
alpha <- 2
beta <- 2
```

a.  theory tells us that $E(X^2) = \frac{\alpha(\alpha + 1)}{(\alpha + \beta + 1)(\alpha + \beta)}$

```{r}
(beta_2nd_moment_theory <- alpha * (alpha + 1) / ((alpha + beta + 1) * (alpha + beta)))
```

b.  deterministic integration

```{r}
beta_integrand <- function(x){
  return(x^2 * dbeta(x, shape1 = alpha, shape2 = beta))
}

(beta_2nd_moment_det <- integrate(beta_integrand, lower = 0, upper = 1))
```

c.  Monte Carlo integration

```{r}
beta_mcarlo_iterations <- 1000

beta_sample <- rbeta(beta_mcarlo_iterations, shape1=alpha, shape2=beta)

(beta_2nd_moment_mc <- mean(beta_sample^2))
```

Monte Carlo errors

```{r}
(beta_mcarlo_error <- sd(beta_sample^2) / sqrt(beta_mcarlo_iterations))
```

Monte Carlo 95% confidence intervals

```{r}
c(
  beta_2nd_moment_mc - 1.96 * beta_mcarlo_error, 
  beta_2nd_moment_mc + 1.96 * beta_mcarlo_error
)
```

Compare the three answers

```{r}
beta_2nd_moment_theory
beta_2nd_moment_det
beta_2nd_moment_mc

c(
  beta_2nd_moment_mc - 1.96 * beta_mcarlo_error, 
  beta_2nd_moment_mc + 1.96 * beta_mcarlo_error
)
```

## Exercise 1

Find the 4th moment of a beta distribution with parameters alpha = 2, beta = 5. Use:

a.  deterministic

b.  Monte Carlo integration

c.  Compute Monte Carlo error and confidence intervals of your estimate(s) from part b.

## Exercise 1 

```{r}
alpha <- 2
beta <- 5
```

a.  deterministic integration

```{r}

```

b.  Monte Carlo integration

```{r}

```

c.  Monte Carlo error and 95% confidence intervals

```{r}

```

Compare the two answers

```{r}

```

## Exercise 2

Integral of $\exp(-x^5)$ on $[0, \infty)$.

Use:

a.  deterministic integration

b.  Monte Carlo integration to approximate it Hint: Uniform sampling won't work, because uniform distribution is not defined on $[0, \infty]$. For importance sampling to work, the support must be the same. Use exponential distribution samples instead -- think about how to write the desired integral in terms of expectation

c.  compute Monte Carlo error and confidence intervals of your estimate(s) from part b

d.  compare Monte Carlo error of the procedure using two different rates of the exponential distribution

## Exercise 2 

Define function

```{r}

```

```{r}

```

Deterministic integration

```{r}

```

Monte Carlo

```{r}

```

Monte Carlo error

```{r}

```

Monte Carlo 95% confidence interval

```{r}

```

Compare to the deterministic numerical integration

```{r}

```

# Importance sampling

Let Z have a standard normal distribution, $Z \sim N(0,1)$. We want to compute $P(X \geq 4.5)$. We could do this via a Naive Monte Carlo simulation. (How?)

```{r}
hist(rnorm(10000,0,1), main = "10,000 Random samples from a Standard Normal")
```

We generate a bunch of samples of Z and count how many satisfy $Z \geq 4$. The problem is that there won’t be very many (perhaps none), unless my number of samples is very large.

**Q:** What is the error of the MC estimate for $p = P(Z \geq 4)$?

Define a threshold value and number of Monte Carlo samples

```{r}
my_const <- 4.5
sim_size <- 1000000
```

To find the true probability of interest:

```{r}
(true_prob <- pnorm(my_const, lower.tail = FALSE))
```

Naive Monte Carlo estimate:

```{r}
naive_samples <- ifelse(rnorm(sim_size) > my_const, yes = 1, no = 0)
(naive_est <- mean(naive_samples))
```

Instead, we can compute this via importance sampling.

We can take the sampling distribution to be a shifted exponential:

$$q(x) = \begin{cases} & e^{-(x-4.5)} \quad \text{when } x \geq 4.5, \\
& 0 \quad \text{ o.w. } \end{cases}$$
(This means that if $Y \sim Exp(1)$, $Y+4 \sim q(x)$)

So the probability that we want to compute, i.e. $p = P(X \geq 4)$

$$p = \int 1_{x \geq 4} \frac{p(x)}{q(x)}q(x)dx$$

Importance sampling estimate:

```{r}
shift_exp <- rexp(sim_size, rate = 1) + my_const
# weights = dnorm(x)/q(x)
is_samples <- dnorm(shift_exp) / exp(-(shift_exp - my_const))
(is_est <- mean(is_samples))
```

```{r}
print(true_prob)
print(naive_est)
print(is_est)
```

Computing Monte Carlo variances

```{r}
(naive_var_mc <- var(naive_samples) / sim_size)
(is_var_mc <- var(is_samples) / sim_size)
```

Computing Monte Carlo confidence intervals

```{r}
# Naive (not very useful, the variance estimate might be degenerate):
c(naive_est - 1.96 * sqrt(naive_var_mc), naive_est + 1.96 * sqrt(naive_var_mc))

# Importance Sampling:
c(is_est - 1.96 * sqrt(is_var_mc), is_est + 1.96 * sqrt(is_var_mc))
```

### Check your understanding

- Try repeating these sames steps with `my_const = 0.9`. Can you explain why the Naive MC performs better in this case?

We will get more samples from the normal distribution in that area.

- Why do you think a shifted exponential is a good proposal distribution in this case?

It generates samples that are in the area of the original distribution that we want to approximate.

- Let's say that you want to use the same shifted exponential (with shift constant = 4), but you want to find $P(Z>8)$. How would you have to modify the Importance Sampling sample to target the correct distribution?

We need to only consider the values where $X>8$ for the I.S. sample (assing 0 to the others).
